\documentclass[aspectratio=169, xcolor=table]{beamer}
\usetheme{TUDelft}

% ====================================================================
% ANIMATION TOGGLE
% ====================================================================
\newif\ifanimated
\animatedtrue
\ifdefined\staticmode
  \animatedfalse
\fi
\newcommand{\cpause}{\ifanimated\pause\fi}

% Import theme macros (includes TikZ libraries and semantic colors)
\input{theme/macros.tex}
\usetikzlibrary{arrows.meta}

% Additional packages for this lecture
\usepackage{booktabs}
\usepackage{listings}
\usepackage[most]{tcolorbox}

% Math definitions
\newcommand{\vq}{\mathbf{q}}
\newcommand{\vp}{\mathbf{p}}
\newcommand{\vd}{\mathbf{d}}
\newcommand{\vk}{\mathbf{k}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\vx}{\mathbf{x}}
\newcommand{\vw}{\mathbf{w}}
\newcommand{\vctx}{\mathbf{c}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

% Legacy color mappings for extracted TikZ
\colorlet{retrieval}{irSparse}
\colorlet{prompt}{irDense}
\colorlet{gen}{irRerank}
\colorlet{ops}{irANN}
\colorlet{warn}{irInfra}

% Title information
\title[RAG Systems 2026]{Designing, Evaluating, Debugging, and Scaling RAG Systems (2026)}
\subtitle{Lecture 7: Beyond ``BM25 + GPT'' --- A Deep Dive into RAG Architecture}
\author{Avishek Anand}
\institute{Information Retrieval}
\date{\today}

\begin{document}

% Title slide
\begin{frame}
\titlepage
\end{frame}

\begin{frame}{Outline (2 hours)}
\small
\begin{enumerate}
  \item \textbf{Framing (15)}: What RAG is (and is not), core axes
  \item \textbf{Retrieval foundation (35)}: retrieval choices, chunking, reformulation
  \item \textbf{Conditioning \& prompting (30)}: context formatting, token economics
  \item \textbf{Reasoning in RAG (30)}: multi-step, ReasonIR/Search-R1 ideas, dynamic search
  \item \textbf{Evaluation (25)}: metrics, failure taxonomy
  \item \textbf{System design \& scaling (25)}: latency, cost, production architecture
\end{enumerate}

\vspace{0.5em}
\begin{tcolorbox}[colback=gray!10,colframe=gray!40,title=Assumed known (not re-taught)]
BM25/RM3/LMs, ANN (IVF+PQ/HNSW), cross-encoders, dense retrieval (DPR/ColBERT/ANCE), hybrid, BEIR/zero-shot issues.
\end{tcolorbox}
\end{frame}

% ==========================================================
\section{Part 0 --- Framing: What is RAG?}

\begin{frame}[plain]
\sectionpage
\end{frame}

\begin{frame}{What RAG Actually Is (and is not)}
\irStage{BACKGROUND}
\textbf{RAG is a compound AI system, not a single model trick or simple recipe.}

\vspace{0.8em}
\begin{columns}[T]
\begin{column}{0.6\textwidth}
\begin{itemize}
    \item \textbf{Not}: Just ``BM25 + GPT'' or simple hallucination reduction.
    \item \textbf{Is}: A pipeline of Retrieval + Conditioning + Generation.
    \item \textbf{Core Axes}: Quality, strategy, reasoning, feedback, and cost.
\end{itemize}
\end{column}
\begin{column}{0.38\textwidth}
\begin{alertblock}{Key Thesis}
\textbf{Retrieval defines the reasoning substrate.}\\
If the evidence is not retrieved, it cannot be used by the model.
\end{alertblock}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{A System View: RAG as a Pipeline of Decisions}
\irStage{SYSTEM ARCHITECTURE}
\textbf{Treat each module as a component with knobs and an evaluation contract.}

\vspace{0.5em}
\begin{center}
\resizebox{0.75\textwidth}{!}{
\input{tikz/rag_system_view.tex}
}
\end{center}

\cpause
\vspace{0.5em}
\begin{block}{Design Principle}
Modularity allows for isolated debugging. If generation fails, we must know if it was due to poor retrieval or poor conditioning.
\end{block}
\end{frame}

\section{Part 1 --- Retrieval: The Substrate (35 min)}

\begin{frame}[plain]
\sectionpage
\end{frame}

\begin{frame}{Retrieval Choices in RAG (2026 View)}
\irStage{RETRIEVAL}
\textbf{Choosing the right retriever depends on the trade-off between precision and OOD robustness.}

\vspace{0.8em}
\begin{table}
\centering
\resizebox{0.9\textwidth}{!}{
\begin{tabular}{l l l}
\toprule
\textbf{Method} & \textbf{Pros} & \textbf{Typical Failure} \\
\midrule
BM25 & Robust OOD, entity precision & Vocabulary mismatch \\
Dense & Semantic matching & Entity drift, OOD issues \\
Hybrid & Best practical default & Complexity, normalization \\
ColBERT & High quality at scale & Storage, infra cost \\
\bottomrule
\end{tabular}
}
\end{table}

\cpause
\vspace{0.2em}
\begin{alertblock}{The Ceiling}
\textbf{Retrieval defines the reasoning substrate.} Most ``hallucinations'' are actually \emph{selection failures}.
\end{alertblock}
\end{frame}

\begin{frame}{Retrieval Substrate: The Evidence Ceiling}
\irStage{THEORY}
\textbf{A grounded answer is a function of the query and the retrieved evidence.}

\vspace{1em}
\begin{columns}
\begin{column}{0.6\textwidth}
\[ \text{Answer}(q) = f\big(q, R_k(q)\big) \]
\cpause
\begin{itemize}
    \item If $e^\star \notin R_k(q)$, then $P(\text{correct} \mid q) \approx 0$.
    \item \textbf{Retrieve}: Find the right chunk.
    \item \textbf{Select}: Include it in context.
    \item \textbf{Use}: Force grounding via prompt.
\end{itemize}
\end{column}
\begin{column}{0.38\textwidth}
\begin{alertblock}{Debugging Mantra}
Before blaming the LLM, verify if the correct evidence was even in the prompt.
\end{alertblock}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Chunking: A Modeling Decision}
\irStage{DATA}
\textbf{Choosing the retrieval unit $u$ impacts recall, precision, and context efficiency.}

\vspace{0.8em}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item \textbf{Extraction}: Tables (linearization), PDFs (layout), metadata injection.
    \item \textbf{Trade-off}: Small chunks $\uparrow$ recall, but $\downarrow$ coherence.
    \item \textbf{Standard}: Fixed-size vs. semantic chunking with overlap.
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\begin{center}
\resizebox{\textwidth}{!}{
\input{tikz/chunking_tradeoff.tex}
}
\end{center}
\end{column}
\end{columns}

\cpause
\vspace{0.5em}
\begin{block}{Rule of Thumb}
Optimize for \textbf{retrievability} and \textbf{citeability}. A chunk without source metadata effectively breaks the RAG attribution chain.
\end{block}
\end{frame}

\begin{frame}{Query Reformulation: Aligning Intent}
\irStage{RETRIEVAL}
\textbf{Bridging the lexical gap between user questions and document structure.}

\vspace{0.8em}
\begin{itemize}
    \item \textbf{HyDE}: Generate a hypothetical document to use as a semantic query.
    \item \textbf{Multi-query}: Generate diverse paraphrases to increase retrieval coverage.
    \item \textbf{Decomposition}: Break multi-part questions into sequential sub-queries.
\end{itemize}

\vspace{1em}
\cpause
\begin{alertblock}{Intent Drift}
Reformulation can shift user intent. Always monitor retrieval overlap between original and rewritten queries to detect drift.
\end{alertblock}
\end{frame}

% ==========================================================
\section{Part 2 --- Conditioning \& Prompting (30 min)}
% ==========================================================
\begin{frame}{How to Feed Retrieved Context}
\irStage{CONDITIONING}
\textbf{Context assembly impacts how well the LLM follows instructions.}

\vspace{0.3em}
\begin{columns}[T]
\begin{column}{0.52\textwidth}
\begin{block}{Context assembly strategies}
\footnotesize
\begin{itemize}
  \item Relevance-sorted + dedup + diversity
  \item Add section headers and provenance
  \item Evidence tags (\texttt{[E1]}, \texttt{[E2]}) for citations
  \item Structured formats: YAML/JSON blocks
  \item Toolformer-style search-as-tool calls
\end{itemize}
\end{block}

\cpause
\begin{alertblock}{Non-negotiable}
\footnotesize
Every chunk must carry provenance: \textbf{doc id, section, timestamp, source}.
\end{alertblock}
\end{column}

\begin{column}{0.45\textwidth}
\footnotesize
\begin{tcolorbox}[colback=gen!5,colframe=gen!50,title=Bad prompt]
\tiny
Here are some documents. Answer the question.
\end{tcolorbox}

\cpause
\vspace{0.2em}
\begin{tcolorbox}[colback=prompt!7,colframe=prompt!60,title=Good prompt]
\tiny
\textbf{Task:} Prevent speculation. Use only evidence.\\
\textbf{Input:} Question + evidence blocks \texttt{[E\#]}.\\
\textbf{Output:} Answer with citations \texttt{[E\#]}.\\
\textbf{Policy:} If evidence insufficient, say ``Insufficient evidence''.
\end{tcolorbox}
\end{column}
\end{columns}
\end{frame}

\begin{frame}[fragile]{A Practical Prompt Template (Evidence-bounded)}
\irStage{CONDITIONING}
\begin{tcolorbox}[colback=gray!8,colframe=gray!40,title=Template, fontupper=\footnotesize]
\begin{lstlisting}[basicstyle=\ttfamily\tiny]
SYSTEM: Use ONLY the evidence blocks. If insufficient, say "Insufficient evidence".
USER QUESTION: {q}
EVIDENCE:
[E1] {source=docA, section=2.1}
{chunk text...}
[E2] {source=docB, section=Table 1}
{chunk text...}
INSTRUCTIONS:
1) Answer in 3-5 bullet points.
2) Every bullet MUST cite an evidence block like [E1].
\end{lstlisting}
\end{tcolorbox}

\cpause
\vspace{-0.5em}
\begin{alertblock}{The Contract}
\footnotesize
Forcing the model to bind claims to evidence reduces hallucinations and enables automated citation verification.
\end{alertblock}
\end{frame}

\begin{frame}{Context Economics: Token Budgeting}
\irStage{SYSTEM DESIGN}
\textbf{The context window is a finite resource that must be managed to maximize reasoning quality.}

\vspace{0.8em}
\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
    \item \textbf{Token Split}: System instructions, evidence, and the generated answer.
    \item \textbf{Recall vs Space}: Adding more chunks $k$ increases recall but reduces per-chunk tokens.
    \item \textbf{Best Practice}: Use a re-ranker to keep only the top-$m$ most relevant chunks.
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\begin{alertblock}{Budgeting Rule}
More context $\neq$ better. Wasted tokens on irrelevant chunks act as noise that degrades model performance.
\end{alertblock}
\end{column}
\end{columns}
\end{frame}

\section{Part 3 --- Reasoning: Beyond Single-Shot}

\begin{frame}[plain]
\sectionpage
\end{frame}

\begin{frame}{Reasoning: Single-shot vs Multi-step}
\irStage{REASONING}
\textbf{Complex queries often require iterative retrieval to resolve multi-hop dependencies.}

\vspace{0.5em}
\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{itemize}
    \item \textbf{Single-shot}: Retrieve once. Fast but fails on ambiguity.
    \item \textbf{Multi-step}: Retrieve $\rightarrow$ reason $\rightarrow$ retrieve.
    \item \textbf{Requirement}: Differentiable stop criteria to control cost/latency.
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{center}
\resizebox{\textwidth}{!}{
\input{tikz/rag_multistep.tex}
}
\end{center}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Retrieval for Reasoning: Chaining Evidence}
\irStage{REASONING}
\textbf{Multi-hop questions require retrieving bridge entities that link $q$ to final $a$.}

\vspace{0.8em}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item \textbf{Constraint}: Evidence $e_1$ must provide information used to find $e_2$.
    \item \textbf{Latent Links}: Bridging entities that are not explicitly in the query.
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\begin{center}
\resizebox{\textwidth}{!}{
\input{tikz/rag_reasoning_chain.tex}
}
\end{center}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Search-in-the-Loop: Dynamic Retrieval}
\irStage{REASONING}
\textbf{Retrieval becomes a tool call within the LLM's reasoning trace (e.g., Search-R1).}

\vspace{0.8em}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item \textbf{Trigger}: Search when model uncertainty $u_t > \theta$.
    \item \textbf{Dynamic}: Fetching new evidence only when internal knowledge is insufficient.
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\begin{center}
\resizebox{\textwidth}{!}{
\input{tikz/rag_dynamic_search.tex}
}
\end{center}
\end{column}
\end{columns}
\end{frame}

% ==========================================================
\section{Part 4 --- Evaluation: The RAG Triad (25 min)}

\begin{frame}[plain]
\sectionpage
\end{frame}

\begin{frame}{RAG Evaluation: Metrics \& Grounding}
\irStage{EVALUATION}
\textbf{Traditional IR metrics (Recall, nDCG) are necessary but insufficient for RAG systems.}

\vspace{0.8em}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item \textbf{Faithfulness}: Is the answer supported by retrieved evidence?
    \item \textbf{Groundedness}: No extra claims beyond the evidence.
    \item \textbf{Attribution}: Do citations point to the correct supporting blocks?
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\begin{alertblock}{The RAG Score}
\small
\[ \text{Score} = w_1 F + w_2 A - w_3 U \]
\tiny $F$: Faithful, $A$: Attrib, $U$: Unsupported.
\footnotesize
Optimizing for correctness without attribution leads to unreliable systems.
\end{alertblock}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Failure Taxonomy: Stop Being Vague}
\irStage{EVALUATION}
\textbf{Design improvements based on specific module-level failure modes.}

\vspace{0.8em}
\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{enumerate}
    \item \textbf{Retrieval}: $e^\star$ is missing from top-$k$.
    \item \textbf{Selection}: $e^\star$ is retrieved but dropped.
    \item \textbf{Reasoning}: $e^\star$ is misused or ignored.
    \item \textbf{Grounding}: Fabricated citations.
\end{enumerate}
\end{column}
\begin{column}{0.5\textwidth}
\begin{center}
\resizebox{\textwidth}{!}{
\input{tikz/rag_failure_taxonomy.tex}
}
\end{center}
\end{column}
\end{columns}
\end{frame}

\section{Part 5 --- System Design: RAG in Production}

\begin{frame}[plain]
\sectionpage
\end{frame}

\begin{frame}{Latency Budgeting: The dominant Term}
\irStage{SYSTEM DESIGN}
\textbf{End-to-end latency is often dominated by generation tokens, not retrieval.}

\vspace{0.8em}
\begin{itemize}
    \item \textbf{Cache}: User queries and top-$k$ results for frequent intents.
    \item \textbf{Parallelize}: Run Sparse and Dense retrieval in parallel.
    \item \textbf{Early Exit}: Stop if retrieval confidence is high after re-ranking.
\end{itemize}

\vspace{1em}
\cpause
\begin{alertblock}{p95 Management}
Optimize p95 with bounded re-ranking. You cannot easily optimize LLM generation speed without reducing output token budget.
\end{alertblock}
\end{frame}

\begin{frame}{Production Architecture: The Real World}
\irStage{INFRASTRUCTURE}
\textbf{Scaling RAG requires decoupled services for retrieval and reasoning.}

\vspace{0.1em}
\begin{center}
\resizebox{0.7\textwidth}{!}{
\input{tikz/rag_production_arch.tex}
}
\end{center}

\vspace{-0.5em}
\cpause
\begin{block}{Hard Truth}
\scriptsize
If you cannot reproduce yesterday's answer by tracing the retrieval logs, you cannot operate a production RAG system.
\end{block}
\end{frame}

\begin{frame}{Summary: The 2026 RAG Engineer}
\irStage{WRAPUP}
\textbf{RAG is a compound system of decisions across data, retrieval, and reasoning.}

\vspace{0.2em}
\begin{itemize}
    \footnotesize
    \item \textbf{Hybrid Default}: Always combine Sparse + Dense.
    \item \textbf{Metadata Chains}: Provenance must be preserved from index to prompt.
    \item \textbf{Citation Contract}: Force the model to bind claims to evidence.
    \item \textbf{Evaluation Logs}: Log retrieve $\rightarrow$ select $\rightarrow$ prompt $\rightarrow$ answer.
\end{itemize}

\vspace{0.2em}
\cpause
\begin{alertblock}{Final Takeaway}
\footnotesize
If you can't trace a claim to evidence, you don't have a RAG system --- you have a text generator.
\end{alertblock}
\end{frame}

\begin{frame}{Questions?}
\centering
\Large \textbf{Thank you!}

\vspace{2em}
\normalsize
Designing, Evaluating, Debugging, and Scaling RAG Systems (2026)
\end{frame}

\end{document}